% Optional: use minted for code listings
%\RequirePackage[]{minted}

\documentclass[
  a4paper,        % select between `a4paper' and `letterpaper'
  techreport,     % select between  `techreport', `report', `article', `commun', `persp' and `review'
  % lineno,         % switch, adds line numbering
  % easyreview,     % use easyReview package
  % anonymous,
  % noauthors,
  % nodate,
]{art}

\input{metadata.tex}
\input{macros.tex}
%\input{templates/pandoc.tex} % needed for pandoc-generated files

\begin{document}
\maketitle

\tableofcontents
\section{Introduction}

% Introduction to actor models and their communication mechanism.
Actor models provide a framework for concurrent systems. In these
systems, the main entity is the notion of an \emph{actor}. Actors are
independent computational units that encapsulate the execution environment of
their processing logic. Asynchronous communication between actors occurs
solely via messages. A message is defined as any piece of data sent from one
actor to another. The reception of these messages is handled by the recipient
actor's \emph{mailbox}, a dedicated piece of storage that preserves message
order while remaining separate from the actor's core logic.

% Challenges in actor system message flow management
One challenge in actor systems is managing message flow under high volumes.
Although some techniques such as rate limiting and \emph{backpressure
routing}~\cite{backpressure} help regulate message rates, these control
mechanisms may also fail under heavily loaded systems. Message delivery is often
assumed to be reliable, but practical implementations must consider the
possibility of message loss and delays from processing bottlenecks or crashing
actors.

% Proposes a solution by decoupling mailboxes from processing actors.
To address some of these challenges, in particular the need to handle message
loss and delays in processing messages, we propose decoupling mailboxes
from \emph{processing actors}, which are the actors that carry out the processing
logic. In this design, communication occurs through message passing to
dedicated \emph{mailbox actors}, with processing actors pulling messages only
when ready to handle them. %
This approach is simple and similar in spirit to Kilim's buffered mailboxes
(multi-producer, single consumer queues) ~\cite{kilim-framework}, allowing us to
give an explicit specification of the message-passing mechanism, where the
mailbox actors remain opaque to both senders and receivers to maintain
compatibility with existing actor frameworks such as Agha's Object-oriented
programming Actor Model~\cite{foundation97agha,taxonomy-of-actor-models}.

% Outlines the benefits of the proposed architectural separation.
The architectural separation of mailboxes from processing actors provides
several key benefits. First, it cleanly separates message management from
processing logic, allowing mailbox actors to specialise in queue management
while processing actors focus on their core functionality. Second, it enables
flexible message scheduling—different mailbox implementations can support
priority queues, batch processing, or custom dispatch strategies without affecting
processing actors. Third, mailbox actors become modular components that can be
modified or replaced independently.

% Discusses the trade-offs and potential drawbacks of the design.
This design may come with some trade-offs. The primary drawback we foresee is
increased system complexity due to additional actors and message-passing
overhead.%
% This manifests as higher resource usage from running more actors and
% potential latency increases when mailbox actors run on different nodes than
% their processing actors.
However, with careful design, these costs can be
outweighed by the benefits of a more flexible system.

% How this paper is organized and what is the contribution.
\paragraph{Contributions}
First, we propose an architectural separation between actors and
their mailboxes in \Cref{sec:model}. This separation enables improved
flexibility and scalability for message management. Second, we formulate
operational semantics in \Cref{sec:semantics} for the decoupled architecture
and its associated explicit message-passing mechanism, formally specifying the
system's design of a variant of Erlang~\cite{formal-semantics-erlang}. This
formal specification provides sufficient grounding to implement a system from
scratch or as a library on top of a compatible system/platform such as Erlang,
and to formalize and prove key system properties such as liveness guarantees and
correctness under concurrent message processing.

\section{Notation}\label{sec:notation}

We use an informal type-theoretic notation. For example, a term $x$ of type $A$
is denoted $\typing{x}{A}$. Capital letters or capitalized words are used for types,
and lowercase letters or lowercase words are used for terms. There is a universe
of types $\U$, so if $A$ is a type, then $\typing{A}{\U}$. We also work with inductive
types presented in the following form:
%
\begin{align*}
  \defequal{\ctype{A}}{\cdots \mid \mathsf{\ccon{C}}(A_1, \dots, A_n) \, ,}
\end{align*}
%
where $\ccon{C}$ is a constructor of the type $\ctype{A}$, and $A_1, \dots, A_n$ are
the types of the constructor arguments.
% If, in the context, there is no ambiguity about the type constructor, we may refer
% to it directly as $\ccon{C}$. Otherwise we use the notation $A.\ccon{C}$ to
% specify the type constructor. This means, type constructor names are not unique,
% and we need to specify the type constructor when we use it.

We use standard type notation where $\N$ denotes the inductive type of natural
numbers, $\EmptyType$ represents the empty type, $\Unit$ denotes the type with
exactly one term denoted with $*$, $\Bool$ denotes the type with two terms,
$\mathsf{tt}$ and $\mathsf{ff}$, and $\List{A}$ represents the type of
parametric lists. Now, $\defequal{x}{y}$ indicates definitional
equality where term $x$ is defined as $y$, while $\equal{a}{b}$ signifies
definitional equality between terms, and $\propequal{a}{b}$ denotes propositional equality.
$A \times B$ denotes the product of $A$ and $B$, and their terms, tuples, are denoted
with $\langle a,b \rangle$, where $a : A$ and $b : B$. %
A \emph{tuple} $\langle a_1, a_2, \dots, a_n \rangle$ is a generalisation of an
ordered pair to an ordered $n$-tuple, where $n \in \N$, of type $A_1 \times
\cdots \times A_n$, and $a_i$ is of type $A_i$ for all $i$ from $1$ to $n$.
Given a type $A$, we can define the optional type $\Option{A}$ as $A \oplus \Unit$.
The terms $\none$ and $\some{a}$ are the
corresponding terms for the optional type when $\typing{a}{A}$.

Some types are sets, and we equip them with some common set-theoretic notations.
The membership relation is denoted with $\in$, so $a \in A$ is a proposition
that $a$ is an element of $A$. Union between sets $A$ and $B$ is denoted with $A
\cup B$. The difference between $A$ and $B$ is denoted with $A \setminus B$.
The disjoint union of $A_1, \dots, A_n$ is denoted with $A_1 \oplus \cdots \oplus A_n$.
%
% and $\mathsf{inj_i}(a_i)$ are the corresponding injections into $A_1 \oplus \cdots
% \oplus A_n$ from $A_i$ for all $i \in \{1, \dots, n\}$.

A function $f$ that takes an argument $a$ of type $A$ and returns a
value of type $B$ is denoted $f : A \to B$ and application is denoted with
$f(a)$. We also use the notation $f[a \mapsto c]$ to define a new function that
is similar to $f$ except that the output of $f$ is replaced by $c$ when the
input is $a$, if $a$ is in the domain of $f$. Otherwise, we extend $f$
accordingly. We overload the notation $\mathcal{N}[a \mapsto c]$ to also apply
to sets, so if $\mathcal{N}$ is a set, then $\mathcal{N}[a \mapsto c]$ is the
set resulting from $\mathcal{N}$ by replacing the element $a$ with $c$, if $a$
is a valid element of $\mathcal{N}$. Otherwise, we extend $\mathcal{N}$
accordingly. Finally, we use (\ignore) to omit terms that are not important or
can be inferred from the context. For example, $\langle \ignore, a, b, \ignore
\rangle$ is a tuple where the first and last elements are irrelevant so the
second and third elements are the only ones that matter.

\section{Model}\label{sec:model}

This section presents a variation of dynamic effective communicating systems
\cite{art-engines} by introducing terminating behaviour guarantees while
preserving their actor-like semantics. Building on this foundation, we formalise
engines as refined effective communicating Moore machines, that is, stateful
message-processing units structured around four core components:
a \emph{configuration}, an \emph{execution environment}, \emph{message interface},
and a \emph{behaviour}. These components, or better their types are used to
define the types for engines for a system, as in~\Cref{def:engine}.

Briefly, the configuration of an engine (\Cref{def:engine-configuration}) conveys  
persistent information about the engine such as its parent, operational mode, and
other engine-type-specific configuration. The engine's execution environment
(\Cref{def:engine-environment}) contains its local state and an \emph{address
book} of other engines to communicate with, including the engine's own address.
The engine's \emph{behaviour} defines how it reacts to messages that conform to
its \emph{message interface} (\Cref{def:message-interface}) ---a sum type of
message types that the engine can process. More precisely, the engine's
behaviour is described through a set of \emph{guarded actions}
(\Cref{def:guarded-action}) consisting of pre-condition/action pairs. When an
engine receives a message, it checks which of these guarded actions have their
pre-conditions satisfied and executes the corresponding actions.
% the following qualifier may not needed after all.
% An action is considered
% \emph{activated} if it belongs to a guarded action whose guard predicate is
% satisfied.

On guarded action evaluation, there are various strategies for selecting which
action to execute when multiple guards are satisfied or have competing priorities.
Although this guard selection problem is beyond the scope of our model, alternative
strategies can be inspired by approaches used in \textsc{Prolog} implementations
and Dijkstra's guarded command language~\cite{dijkstra-guarded-commands}. For
demonstration purposes, we present one such strategy in \Cref{sec:behavior-eval-rules}.
However, this is not essential to the model presented here, and readers are encouraged
to define alternative strategies according to their needs.

\begin{remark}[Event-Driven Systems]\label{rem:event-driven-systems}%
  The behaviour of engines, as described above, resembles the correct-by-construction
  approach used in event-driven systems such as Event-B~\cite{dupont:tel-04165215}.
  In this approach, system transitions are driven by the \emph{events} that trigger
  them. An event primarily consists of a set of preconditions on the current state
  of the machine and a set of actions to be executed when these preconditions are
  all satisfied. The preconditions are typically expressed in non-deterministic
  first-order logic, and the actions are defined using a language derived from
  process calculi, which often includes operations such as parallel composition,
  a null operation, sequencing, nondeterministic choice, and recursion.
  %Here, we have a more restricted set of actions, as will be seen in \Cref{def:effects}.
  % non-deterministic first-order logic, and the actions are defined using a
  % language derived from process calculi, which often includes operations such as
  % parallel composition, a null operation, sequencing, nondeterministic choice,
  % and recursion.
  % Here, we have a more restricted set of actions, as will be seen in \Cref{def:effects}.
\end{remark}

% Object model for distributed systems (investigate)

\subsection{Preliminaries}\label{sec:preliminaries}%

The model operates on a system state $\SystemState$, consisting of finite sets
of uniquely identified nodes $\SetOfNodes$, engines $\SetOfEngines$, and
in-transit messages $\SetOfMessages$. The subsequent sections detail these
concepts in a bottom-up manner.

As a state machine, the system evolves in discrete time steps, where the state
at step $i$ is denoted with $\SystemState^{(i)}$. The sets $\SetOfNodes^{(i)}$,
$\SetOfEngines^{(i)}$, and $\SetOfMessages^{(i)}$ may also change between steps.
When no confusion arises, we omit the step index for any of these entities,
including the system state. Each engine resides on a unique node and its type is
as defined in \Cref{def:engine,def:node}, respectively.

\begin{remark}[Static Engine Allocation]\label{rem:move-engines}%
  A system might relocate an engine to a different node when computational costs
  in its current node become prohibitive and more favourable resource
  availability exists elsewhere. This, is however, not considered in this model.
  Here, the model assumes that engines always remain on the same node where they
  were initially created.
\end{remark}

The communication between engines is done via messages, which are typed and
carried by the system. A message-passing mechanism governs the delivery of
in-transit messages $\SetOfMessages$, see \Cref{sec:message-structure}. Each
message contains sender and recipient identifiers along with a payload.
Similarly, as engine types, message types are fixed once the system is defined.

\begin{remark}[Static Typing]\label{rem:static-typing}%
  This model employs static typing. Engine types are predetermined with the
  system, and new engine instances must conform to these types. The sum type
  of these engine types is denoted with $\mathsf{Eng}$. Similarly, message
  payloads adhere to predefined types specific to the target engine.
  The sum type of these message types is denoted with $\Msg$.
  Extending this model to support dynamic typing is future work.
\end{remark}

% \begin{itemize}
%   \item Addresses and names in \Cref{sec:addresses-and-names}
%   \item Message structure in \Cref{sec:message-structure}
%   \item Engine components in \Cref{sec:engine-components}
%   \item Execution directives in \Cref{sec:execution-directives}
% \end{itemize}

\subsection{Addresses}\label{sec:addresses-and-names}

The system needs a way to identify each engine uniquely. This is done by
combining the host node's identifier with a local engine identifier that is
unique within that node. For simplicity, we use natural numbers to represent
both the node and the engine identifiers.

\begin{definition}[Address]%
\label{def:address}
The \textbf{address} of an engine is a tuple $\loc{n}{\iota}$ of type
$\Address$ where:
\begin{itemize}
    \item $n$ is the identifier of the node, and
    \item $\iota$ is the engine's unique identifier within the node.
\end{itemize}


% \begin{notation}\label{notation:address-notation}%
% If $t$ is an address, to reference the engine with that
% address, we use the notation $\engine{t}$. In case of nodes, we use the notation
% $\node{i}$ to refer to the node with identifier $i$.
% \end{notation}
\end{definition}

% We define projection functions $\mathsf{node}(a)$ and $\uid{a}$ to
% extract the respective components of an address $a$.

\subsection{Message structure}\label{sec:message-structure}

Engines communicate via typed messages. The system uses a static typing approach
where message types are predefined for each engine.

\begin{definition}[Message Type]\label{def:message-type}%
A \textbf{message type} is a type that within the system is used as the type
for payloads of messages that can be processed by engines.
\end{definition}

For example, $\Bool$, $\N$, or $\List{\N}$ could be message types.

\begin{definition}[Message Interface]\label{def:message-interface}%
The \textbf{message interface} of an engine $p$ is a sum type where its type
constructors, called them \textbf{message tags}, wrap a message type with a given
semantic meaning that the engine can process.
\end{definition}

As our running example, we will consider the engine type $\mathsf{kv}$ that
can process messages of type in \eqref{eq:message-interface-example}. The complete
engine type $\mathsf{kv}$ would be defined later, once all required constructs are
presented.

% The following is our running example of an engine.

\begin{example}[Key-Value Store Engine]\label{ex:keyvalue-engine}%
  Let us assume the system defines the engine type $\enginetype{\mathsf{kv}}$
  and that an engine of this type works as a simple key-value store. An engine
  of this type has the following message interface:
  \begin{equation}\label{eq:message-interface-example}
  \typedef%
  {\msgtype{\mathsf{kv}}}%
  {    \msg{Put}(\mathsf{Key}\times\mathsf{Val}) 
  \mid \msg{Get}(\mathsf{Key}) 
  \mid \msg{Delete}(\mathsf{Key})
  \mid \msg{Result}(\Option{\mathsf{Val}})
  },
\end{equation}

where $\msg{Put}$, $\msg{Get}$, and $\msg{Delete}$ are message tags
representing the operations that the key-value store engine can perform to
modify or query its memory. The message type $\mathsf{Key}\times\mathsf{Val}$
that $\msg{Put}$ wraps encodes a key-value pair, representing the data to be
stored. The message type $\mathsf{Key}$ that $\msg{Get}$ and $\msg{Delete}$ wrap
represents the key to retrieve or remove. The message tag $\msg{Result}$ wraps
an optional value, representing the result of a query operation which may or may
not find a value for the requested key.
\end{example}

We model the system in such a way that it enjoys type-safe communication. This
is possible because of how we define a message.

\begin{definition}[Message]\label{def:messages}
A \textbf{message} is a triple $\langle s, t, v \rangle$ 
denoted with $\msgpacket{s}{t}{v}$
where:
\begin{itemize}
    \item $\typing{s}{\Address}$ is the sender engine address,
    \item $\typing{t}{\Address}$ is the target engine address, and
    \item $\typing{v}{\msgtype{t}}$ is the payload conforming to the target
    engine's message interface.
\end{itemize}
% We define projection functions $\sender{m}$, $\target{m}$, and
% $\mathsf{msg}(m)$ to extract the respective components of a message.
\end{definition}

\begin{example}[Message Example]\label{ex:message-example}%
  A message sent to the key-value store engine might be 
  $\msgpacket{\loc{1}{2}}{\loc{3}{4}}{\msg{Put}(\text{cd2edf}, \text{Alice})}$
  where the sender resides on node $1$ with engine identifier $2$, and the target
  resides on node $3$ with engine identifier $4$. The payload is a key-value pair,
  conforming to the message interface $\msgtype{\mathsf{kv}}$.
\end{example}

To refer to an arbitrary type of a message \emph{within the system} at any step, we use the type
$\Msg$, the sum type of all message interfaces, and therefore the type of all
valid messages, defined as,
\begin{equation}\label{eq:message-type-sum}
\defequal{\Msg}{\bigoplus_{i} \msgtype{i}},
\end{equation}
where $\msgtype{i}$ is the message interface of engine type $i$. To simplify the
notation, we assume message type subtyping, that is, $\typing{\msgtype{i}}{\Msg}$
for all $i \in \SetOfEngines$. When no confusion arises, using $p$ to refer to an
engine, we use $\msgtype{p}$ to refer to its message interface. 
% Also, we may use $\enginetype{p}$ to refer to the engine type of $p$.

% Ill-formed message example
% \begin{example}[Ill-formed Message]\label{ex:ill-formed-message}%
%   The message $\msgpacket{s}{t}{\msg{Add}(1, 2, 3)}$ sent to the calculator
%   engine is ill-formed because the payload $\msg{Add}(1, 2, 3)$ does not conform
%   to the message interface $\msgtype{\mathsf{calc}}$, which expects exactly two
%   natural numbers as arguments. Another example of an ill-formed message is
%   $\msgpacket{s}{t}{\msg{Multiply}(\mathsf{tt}, 2)}$ as it expects a pair of
%   natural numbers as payload, but receives a boolean value $\mathsf{tt}$ as the
%   first argument instead.
% \end{example}

% \begin{remark}[Untagged Unions]\label{rem:untagged-unions}%
% An alternative approach defines message interfaces through untagged unions, permitting
% overlapping message types across engines. This is consistent with our approach,
% enabling shared message constructors between engines capable of processing them,
% though it diverges from our system's concrete implementation in Juvix.
% \end{remark}

\subsection{Engine components}\label{sec:engine-components}

In the system, an engine is a term of type $\enginetype{i}$ where $i$ is one of
the predefined engine types. These engines are composed of various components,
some containing engine-specific data that remains unchanged throughout the
engine's lifecycle such as its parent reference, while others may change during
execution such as the engine's local state.

\begin{definition}[Engine Lifecycle]\label{def:engine-status}%
The \textbf{status} of an engine $p$ of type $\EngineStatus{p}$
determines the engine's operational state within its lifecycle as shown in
\Cref{fig:engine-status}.

\begin{equation}\label{eq:engine-status}
\begin{aligned}
\typedef%
  {\EngineStatus{p}}%
  {    &\quad \ready{\Option{\msgtype{p} \to \Bool}} \\%
  \mid & \quad \busy{\msgtype{p}} \\%
  \mid & \quad \terminated  \, .
  }
\end{aligned}
\end{equation}

% tikz
\begin{figure}%[!hbp]
\centering
\[\begin{tikzcd}[row sep=1em, column sep=2em]
	& {\textcolor{darkgreen}{\ready{f}}} \\
	{\textcolor{darkred}{\terminated}} && {\textcolor{darkorange}{\busy{m}}}
	\arrow[from=1-2, to=2-1]
	\arrow[from=1-2, to=2-3, bend right]
	\arrow[from=2-3, to=1-2, bend right]
	\arrow[from=2-3, to=2-1, bend left]
\end{tikzcd}\]%
\caption{The engine life cycle: every engine starts in \ready{f} state waiting
for a message where $f$ is an optional predicate that communicates which
messages the engine is interested in processing. Once it reacts to a message, it
changes to \busy{m} where $m$ is the message to which the engine reacts. 
Then, depending on its logic, the engine may either return to \ready{f'} for
some optional predicate $f'$ or \terminated{}, meaning the engine will not process
any more messages.}
\label{fig:engine-status}
\end{figure}
\end{definition}

As illustrated in \Cref{fig:engine-status}, an engine in the \ready{f} state can
process messages. The optional predicate $f$ specifies which message types the
engine wants to receive. When no predicate is provided (\none), all messages
pass through by default. This is equivalent to having a predicate that always
returns $\mathsf{tt}$. At the creation time, unless otherwise specified, an engine
is in the \ready{\none} state, meaning it accepts all messages by default.

After creation, the engine may transition to the \busy{m} state, where it is
actively processing message $m$, or to the \terminated{} state, where it stops
processing messages and is removed from the system in the next step.

\begin{example}\label{ex:engine-status-example}%
  A key-value store engine as in \Cref{ex:keyvalue-engine} may start in the
  \ready{\some{\mathsf{f}}} state, where $\mathsf{f}$ is the filter that rejects
  \msg{Delete} messages, as in \eqref{eq:engine-status-example}. This allows the
  store to be safely populated before any deletion operations are permitted. Once
  the store is populated, the filter can be updated, for example, to \none{} so
  that all messages are accepted.
  \begin{equation}\label{eq:engine-status-example}
    \mathsf{f} := \lambda m.%
  \mathsf{case}\ m\ \mathsf{of}\ \{%
    \msg{Delete}(x) \Rightarrow \mathsf{ff};%
    \ignore \Rightarrow \mathsf{tt}%
    \},
  \end{equation}
\end{example}

% \begin{remark}[Engine Removal]\label{rem:engine-removal}%
%   Even with an engine in the \terminated{} state, its address may still be
%   referenced in messages. There is no way to prevent this effectively.
%   % Since the system does not maintain a registry of terminated engines, nor
%   % does it notify senders when they attempt to communicate with non-existent
%   % engines. One might want to extend the system so that it can notify senders
%   % when they attempt to communicate with non-existent engines.
% \end{remark}

We now introduce the notion of \emph{operational mode} to distinguish between
two types of engines, processing engines and mailbox engines. This distinction is
based on the engine's purpose and how messages are routed to it.

\begin{definition}\label{def:engine-mode}
The \textbf{operational mode} of an engine is of type $\EngineMode$,
\begin{equation}\label{eq:engine-mode}
\typedef{\EngineMode}
  {    \process 
  \mid \mail
  }\,.
\end{equation}
\end{definition}

The operational mode of an engine cannot change after engine creation. In fact,
this mode is opaque to the engine itself and is defined by the system. The
system's message-passing mechanism treats engines differently based on their
mode. In short, \process engines do not receive messages directly. Messages sent
to them are internally routed to \mail engines, which can accept all messages
and route them back to their \process engine when appropriate. The operational
mode is stored in the engine's configuration.

\begin{definition}\label{def:engine-configuration}%
Given a type $C_p$, the \textbf{configuration} of an engine $p$ parameterised by $C_p$ is
a tuple $\configval{p}{\mode}{c}$ where:
\begin{itemize}
    \item $\typing{p}{\Option{\Address}}$ is an optional reference (address) of the engine's parent,
    \item $\typing{\mode}{\EngineMode}$ is the engine's \hyperref[def:engine-mode]{operational mode}, and
    \item $\typing{c}{C_p}$ is configuration data specific to the engine type of $p$.
\end{itemize}
The type of engine configurations for engine type $\enginetype{p}$ is denoted
with $\configtype{p}{C_p}$. 
% The type $S_p$ comes from the engine's type.
Given an engine $p$, we use $\config{p}$ to refer to its configuration.
\end{definition}

\begin{example}\label{ex:keyvalue-store-configuration}%
Let us define the engine-specific configuration type $C_p$ as $\ctype{C}_{\mathsf{kv}}$
for the key-value store engine from \Cref{ex:keyvalue-engine}. This type
establishes the operational constraints of the key-value store, similar to how
database systems often have different access/update modes.
\begin{equation}\label{eq:keyvalue-store-configuration}
\begin{aligned}
\typedef{%
  \ctype{C}_{\mathsf{kv}}
}{      &\quad \ccon{ReadOnly} & \text{(only allows reading existing entries)} \\
   \mid & \quad \ccon{ReadAppend} & \text{(allows reading and adding new entries)} \\
   \mid & \quad \ccon{ReadModify} & \text{(allows reading, adding, and updating entries)} \\
   \mid & \quad \ccon{ReadWrite} & \text{(full CRUD capabilities)}
}.
\end{aligned}
\end{equation}
For convenience, given a key-value store engine $p$, and its configuration
$\config{p}$, we define predicates that check if the configuration is of a
certain mode, $\mathsf{isReadOnly}$, $\mathsf{isReadAppend}$,
$\mathsf{isReadModify}$, and $\mathsf{isReadWrite}$. 
\end{example}

While engines operate in isolation and process messages independently, their
configuration alone cannot support their full functionality. Each engine
requires additional state management capabilities to operate effectively.
To address this need, every engine maintains its own state within an execution
context, collectively referred to as the engine's \emph{environment}. 

\begin{definition}\label{def:engine-environment}%
Given a type $L_p$ for the local state of an engine $p$ as defined by its engine
type, the execution \textbf{environment} of an engine $p$ is a tuple
$\localenvval{s}{m}$, of type $\envtype{p}{L_p}$, where:
\begin{itemize}
    \item $\typing{s}{L_p}$ is the local state of the engine, and
    \item $\typing{m}{\List{\Address}}$ is the engine's address book.
\end{itemize}
Given an engine $p$, we use $\localenv{p}$ to refer to its environment.
\end{definition}

\begin{remark}\label{rem:engine-environment}%
Every engine's address book contains at least its own address, which is
automatically added during engine spawning.
% For convenience, the system also provides a function \self{} that
% refers to the engine's own address.
\end{remark}
\begin{example}\label{ex:keyvalue-store-engine-local-state}%
For the key-value store engine, we define the type $L_{\mathsf{kv}}$ for the
local state to include both the store and access statistics:

\begin{equation}\label{eq:keyvalue-store-engine-local-state}
\defequal{L_{\mathsf{kv}}}{
          ({\mathsf{Key}}\to {\Option{\mathsf{Val}}})
   \times ({\mathsf{Key}}\to {\N})
   \,.
}
\end{equation}
Therefore, we can use as initial local state of a key-value store engine
the pair,
\begin{equation}\label{eq:keyvalue-store-engine-initial-local-state}
\defequal{
  \typing{
    l_{\mathsf{kv}}
  }{ 
    L_{\mathsf{kv}}
  }
  }{
    (\lambda k.\ \none, \lambda k.\ 0)
  }
\,,
\end{equation}
so, it represents an empty store, returning $\none$ for any key, and an access
counter initialised to zero for all keys. For convenience, given a key-value
store engine $p$ with environment $\localenv{p}$, we define the projections
$\mathsf{store}$ and $\mathsf{accessCount}$ to access these components.
\end{example}

\subsection{Execution directives}\label{sec:execution-directives}

The behaviour of an engine specifies what it can perform when reacting to a
message. How the engine reacts to a message is determined by a set of guarded
actions ---pairs of pre-conditions called \emph{guards} and their corresponding
\emph{actions}. When a message is being processed, the system evaluates it
against the engine's guards, considering both the engine's environment and
configuration. If a guard is satisfied, its associated action produces an
\emph{effect} that the system executes, as illustrated in \Cref{fig:effect}.

\begin{definition}[Effect System]\label{def:effects}%
Let $C_p$ be the type representing the engine-specific configuration of an
engine~$p$ as in \Cref{def:engine-configuration}, and $L_p$ be the type representing
the local state of an engine~$p$ as in \Cref{def:engine-environment}. An
\textbf{effect} of type $\effecttype{p}{C_p}{L_p}$ represents a primitive
operation that an engine can perform to influence the system state when
processing a message.

\begin{equation}\label{eq:effects}
\begin{aligned}
\typedef{%
   \effecttype{&p}{C_p}{L_p}
}{\\
      &\quad \noop & &%
        \text{(null operation)} \\
\mid & \quad \send{\Address}{\Msg} & &%
        \text{(message dispatch)} \\
\mid & \quad \update{\envtype{p}{L_p}} & &%
        \text{(env. modification)} \\
\mid & \quad \spawn{\configtype{p}{C_p}}{\envtype{p}{L_p}} & &%
        \text{(engine creation)} \\
\mid & \quad \mfilter{%
                \Option{\msgtype{p} \to \Bool}
              }& &%
        \text{(new mailbox filter)} \\
\mid & \quad \terminate & &%
        \text{(engine shutdown)} \\
\mid & \quad \chain{\effecttype{p}{C_p}{L_p}}{\effecttype{p}{C_p}{L_p}} & &%
        \text{(action sequencing)\, .}
}
\end{aligned}
\end{equation}
\end{definition}

% TODO: mention how this induces a monad.
% also mention CCS?
\begin{figure}%[!htbp]
% https://q.uiver.app/#q=WzAsOSxbMSwwLCIobSwgXFxtYXRoc2Z7Y2ZnX2p9LFxcbWF0aHNme2Vudn1faikiXSxbMCwyLCJyXzEiXSxbMSwyLCJcXGNkb3RzIl0sWzIsMiwicl9rIl0sWzAsMywiZV8xOlxcbWF0aHNme0VmZmVjdH0oU19qKSJdLFsyLDMsImVfazpcXG1hdGhzZntFZmZlY3R9KFNfaikiXSxbMCwwXSxbMiwwXSxbMSwzLCJcXGNkb3RzIl0sWzAsMiwiZ19pIiwyLHsic3R5bGUiOnsidGFpbCI6eyJuYW1lIjoibWFwcyB0byJ9fX1dLFswLDMsImdfayIsMCx7InN0eWxlIjp7InRhaWwiOnsibmFtZSI6Im1hcHMgdG8ifX19XSxbMSw0LCJjXzEiLDIseyJzdHlsZSI6eyJ0YWlsIjp7Im5hbWUiOiJtYXBzIHRvIn19fV0sWzMsNSwiY19rIiwwLHsic3R5bGUiOnsidGFpbCI6eyJuYW1lIjoibWFwcyB0byJ9fX1dLFswLDEsImdfMSIsMix7InN0eWxlIjp7InRhaWwiOnsibmFtZSI6Im1hcHMgdG8ifX19XSxbNiw3LCIiLDAseyJsYWJlbF9wb3NpdGlvbiI6MTAwLCJjdXJ2ZSI6NCwic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZGFzaGVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiw4LCIiLDIseyJzdHlsZSI6eyJ0YWlsIjp7Im5hbWUiOiJtYXBzIHRvIn19fV1d
\[\begin{tikzcd}
	{} & {(m, \config{p},\localenv{p})} & {} \\
	\\
	{r_1} & \cdots & {r_k} \\
	{e_1:\effecttype{p}{C_p}{L_p}} & \cdots & {e_k:\effecttype{p}{C_p}{L_p}}
	\arrow[curve={height=24pt}, shorten <=24pt, shorten >=24pt, dashed, no head, from=1-1, to=1-3]
	\arrow["{g_1}"', maps to, from=1-2, to=3-1]
	\arrow["{g_i}"', maps to, from=1-2, to=3-2]
	\arrow["{g_k}", maps to, from=1-2, to=3-3]
	\arrow["{c_1}"', maps to, from=3-1, to=4-1]
	\arrow[maps to, from=3-2, to=4-2]
	\arrow["{c_k}", maps to, from=3-3, to=4-3]
\end{tikzcd}\]%
\caption{For an engine $p$, its behaviour acts on a message $m$ of type
$\msgtype{p}$ and an engine's configuration $\config{p}$
and environment $\localenv{p}$. The pre-conditions, denoted as $g_i$, are
guards that determine which action is executed. The symbols $r_i$ represent
output that feeds into the actions $c_i$. The actions, denoted as $c_i$ return
one of the effects as in \Cref{def:effects}. These effects are executed if their
pre-conditions are satisfied. If no pre-conditions are satisfied, the action is a
null operation, $\noop$.}
\label{fig:effect}
\end{figure}


\begin{definition}[Guard]\label{def:guards}%
Under the same assumptions as in \Cref{def:effects}, given an engine $p$, and a
type $\returntype$, a \textbf{guard} is a mapping of type
\begin{equation}
\In{p} \to \Option{\returntype} \text{ where }
\defequal{\In{p}}{\msgtype{p} \times \configtype{p}{C_p} \times \envtype{p}{L_p}}\,.
\end{equation}
It returns a value of type $\returntype$ if the guard is satisfied, and
$\none$ otherwise. Recall that $L_p$ is the type of the engine's local state.
\end{definition}

As illustrated in Figure~\ref{fig:effect}, guards act as pre-conditions that
determine which actions should be executed. We can define a corresponding
predicate for any guard $g$, where the predicate $\hat{g}$ maps to $\mathsf{ff}$
if and only if $g$ maps to $\none$, otherwise it maps to $\mathsf{tt}$. When a
guard is satisfied, its associated action is triggered, leading to the execution
of specific effects by the system's runtime.

% On the computational aspects of an engine, the behaviour evaluating the
% different guards, and for each guard, the action to be executed, whenever the
% guard is satisfied, is as follows:

\begin{definition}\label{def:action}%
Similarly, as in the case of guards, an \textbf{action} acting on data of type
$\returntype$ is a mapping of type %
\begin{equation}
\returntype \times 
\In{p}
  \to 
\Out{p} \text{ where }
\defequal{\Out{p}}{\effecttype{p}{C_p}{L_p}}.
\end{equation}
\end{definition}

\begin{definition}[Guarded action]\label{def:guarded-action}%
  Under the same assumptions as in \Cref{def:guards,def:action}, given an engine
  $p$, a \textbf{guarded action} for engine $p$ is a tuple
  $\guardedaction{i}$ where:
  \begin{itemize}
    \item $\typing{\guard{i}}{\In{p} \to \Option{\returntype}}$ is a guard that determines when the
    action $\action{i}$ should execute, and
    \item $\typing{\action{i}}{\returntype \times \In{p} \to \Out{p}}$ is an action that specifies
    an effect.
  \end{itemize}
  Together, both, the guard and action induce a mapping of type $\In{p} \to
  \Out{p}$ that takes in a message, engine configuration, and environment, and
  outputs the effect to execute when the guard is satisfied.
\end{definition}

\begin{example}\label{ex:keyvalue-engine}%
For the key-value store engine, we define guarded actions for each message type.
The environment and configuration types are
$\envtype{\mathsf{kv}}{L_{\mathsf{kv}}}$ and
$\configtype{\mathsf{kv}}{\ctype{C}_{\mathsf{kv}}}$ as in
\Cref{ex:keyvalue-store-engine-local-state,ex:keyvalue-store-configuration}. As
shown in \eqref{eq:keyvalue-engine-get-guard}, we can define the guard
$g_{\mathsf{get}}$ to check for the $\msg{Get}$ tag and return the key if found,
otherwise $\none$. The action $a_{\mathsf{get}}$ retrieves the value, updates
the access count, and sends the result back. We assume that the sender engine,
the consumer of the key-value store engine's result, also has
$\msg{Result}(\mathsf{Val})$ and $\msg{Ack}$ as part of its message interface,
so we can write back the result directly. 

\begin{equation}\label{eq:keyvalue-engine-get-guard}
  \begin{minipage}{0.45\textwidth}
  \begin{align*}
  &\typing{
    g_{\mathsf{get}}
  }{
    \In{\mathsf{kv}} 
      \to 
    \Option{\mathsf{Key}}
    \,.
  } \\
  &\defequal{
    g_{\mathsf{get}}(\langle m, \ignore, \ignore \rangle)
  }{
    \mathsf{case}\ m\ \mathsf{of}\ \{\\
      &\hspace{3mm} \msg{Get}(k) \Rightarrow \some{k};\\
      &\hspace{3mm} \ignore \Rightarrow \none
    \}
    \,.
  }
  \end{align*}
  \vfill
  \end{minipage}
  \hfill\vrule width 1pt\quad\hfill
  \begin{minipage}{0.48\textwidth}
  \begin{align*}
  &\typing{
    a_{\mathsf{get}}
  }{
    \mathsf{Key} \times \In{\mathsf{kv}} 
      \to 
     \Out{\mathsf{kv}}
    \,.
  } \\
  &\defequal{
    a_{\mathsf{get}}(k, \langle m, c, \langle \mathsf{st}, \mathsf{cnt} \rangle \rangle)
  }{ \chain{\\
  &\quad  \update{%
        \langle
          \mathsf{st},
          \mathsf{cnt}[k \mapsto \mathsf{cnt}[k] + 1]
        \rangle
      }
      }{ \\
  &\quad \send{\sender{m}}{\msg{Result}(\mathsf{st}[k])}
      }
  }\,.
  \end{align*}
  \vfill
  \end{minipage}
\end{equation}

Similarly, for $\msg{Put}$ messages, we define a guard that checks if the engine
is not in read-only mode, and an action that updates the store and acknowledges
the sender.

\begin{equation}\label{eq:keyvalue-engine-put-guard}
  \begin{minipage}{0.38\textwidth}
  \begin{align*}
  &\typing{
    g_{\mathsf{put}}
  }{
    \In{\mathsf{kv}} 
      \to 
    \Option{\mathsf{Key} \times \mathsf{Val}}
    \,.
  } \\
  &\defequal{
    g_{\mathsf{put}}(\langle m, c, \ignore \rangle)
  }{\\
    &\quad \mathsf{if}\,(\neg\,\mathsf{isReadOnly}(c))\\\
    &\quad \mathsf{then} \{ \mathsf{case}\ m\ \mathsf{of}\ \\
    &\quad \quad \big \{ 
    \begin{array}{l}
    \msg{Put}(\langle k, v \rangle) \Rightarrow \some{\langle k, v \rangle};\\
    \ignore \Rightarrow \none\\
    \end{array}
    \big \}\\
    &\quad \} \mathsf{else}\ \none
    \,.
  }
  \end{align*}
  \vfill
  \end{minipage}
  \hfill\quad \vrule width 1pt\quad\hfill
  \begin{minipage}{0.40\textwidth}
  \begin{align*}
  &\typing{
    a_{\mathsf{put}}
  }{
    (\mathsf{Key} \times \mathsf{Val}) \times \In{\mathsf{kv}} 
      \to \Out{\mathsf{kv}}
    \,.
  } \\
  &\defequal{
    a_{\mathsf{put}}(\langle k, v \rangle, \langle m, c, \langle \mathsf{st}, \mathsf{cnt} \rangle \rangle)
  }{ \\
  &\quad \chain{\\
  &\quad \quad      \update{%
        \langle
          \mathsf{st}[k \mapsto v],
          \mathsf{cnt}
        \rangle
      }
      }{ \\
  &\quad \quad  \send{\sender{m}}{\msg{Ack}}
      }
  }\,.
  \end{align*}
  \vfill
  \end{minipage}
\end{equation}
\end{example}

%TODO: comment on the given guard action evaluation strategy in the corresponding op. rule

\begin{remark}\label{rem:turing-completeness}%
A question at this point is whether engines can enable unbounded computation.
Behaviours as introduced in~\cite{art-engines} are Turing complete. To see this,
we need to show that the design satisfies the two fundamental requirements:
conditional branching through guard-state interactions and indefinite looping
via message-passing recursion. Note that the message-passing mechanism says
nothing about forbidding engines from sending messages to themselves, which is a
powerful mechanism that allows for recursive patterns. So, while individual
effects as listed in \Cref{def:effects} lack explicit looping constructs, and it
may not seem there is a recursive mechanism at first glance, the combination of
guarded actions acting as conditional statements with open-ended message passing
enables computational universality.
\end{remark}

\begin{definition}[Engine behaviour]\label{def:engine-behaviour}
Given an engine $p$, the type of its behaviour is denoted with
$\behaviourtype{p}$, and it is the set of its guarded actions.
\end{definition}

\subsection{Engines}\label{sec:processing-engines}

Having defined the components that make up an engine's behaviour, we can now
formally specify engines themselves. As the fundamental computational units of
our system, engines encapsulate both the processing logic and state management
capabilities. We have established the types of their constituent parts, that is,
guards, actions, configurations and environments.

\begin{definition}[Engine]\label{def:engine}%
An \textbf{engine} is the computational unit of the system, formally specified
by a tuple $\engineval{i}{b}{s}{\mathsf{cfg}}{\mathsf{env}}$ where:
\begin{itemize}
\item $i$ is the engine's message interface, a type as described in \Cref{def:message-interface},
\item $b$ is the engine's behaviour ---a set of guarded actions as defined in \Cref{def:guarded-action},
\item $s$ is the engine's status flag as specified in \Cref{def:engine-status},
\item $\mathsf{cfg}$ is the engine's configuration as described in
\Cref{def:engine-configuration}, and
\item $\mathsf{env}$ is the engine's execution environment as outlined in \Cref{def:engine-environment}.
\end{itemize}
\end{definition}

% \begin{remark}\label{rem:engine-init}
% No need to initialize an engine. Once declared, an engine is ready to process
% messages.
% \end{remark}

Our system defines two engine types: \emph{processing engines} and \emph{mailbox
engines}. Users can only explicitly create processing engines, while mailbox
engines are automatically created by the system when a processing engine is
instantiated, as detailed in \Cref{sec:creation-rules}. A processing engine is
distinguished by having its operational mode set to $\process$ through
its configuration's mode field. The internals of mailbox engines are described
in \Cref{def:mailbox-engine}.

\begin{definition}[Mailbox engine]\label{def:mailbox-engine}
A \emph{mailbox engine} $\mailbox{j}$ is a system-pre-configured engine
automatically created when instantiating a processing engine $\engine{j}$. It must
satisfy:
\begin{itemize}
\item Operational mode set to $\mail$ (per \Cref{def:engine-mode})
\item Parent reference pointing to its paired processing engine $\engine{j}$
\item Fixed message interface with one message type:
\begin{equation*}
  \typedef%
    {\msgtype{\mailbox{j}}}
    {\append(\msgtype{j})}
\end{equation*}
carrying a payload of type $\msgtype{j}$
\item State $q$ of type $S$ implementing a queue-like structure with:
  \begin{itemize}
  \item $\emptyset$ denoting empty state.
  \item Message insertion to support $\append\mbox{-}$messages, %
  $\typing{\insertmsgop}{S \times \msgtype{j} \to S}.$
  \item Message deletion, %
  $\typing{\deletemsgop}{S \times \msgtype{j} \to S}.$
  % for verification, one might want to add proof that the operations
  % actually insert/delete the element from the structure
  \end{itemize}
\end{itemize}
\end{definition}

\noindent With mailbox engines added to the system, we have a strict producer-consumer
pattern introduced through explicit message-reception semantics, decoupling
message storage from processing logic. This allows for independent scaling of
message ingestion and core computation.
%TODO: add a reference to the corresponing rules in the operational semantics

\subsection{Nodes}\label{sec:nodes}

Having previously introduced the system $\SystemState$, which comprises a set of nodes%
~$\SetOfNodes$ that host engines $\SetOfEngines$ governed by a message-passing
mechanism defined by the system's requirements, we now formally define what
constitutes a node.

\begin{definition}[Node]\label{def:node}%
A \textbf{node} in $\SetOfNodes$ is a tuple
$\nodeval{n_i}{\enginemap_i}{\plugins_i}$ where $\node{i}$ corresponds to the
node's unique identifier across the system, $\enginemap_i$ is a mapping from
identifiers to engine instances on the node, and $\plugins_i$ represents
node-specific functionality, including, for example, storage and freshness of
unique engine identifiers.
\end{definition}

For convenience, given a node with identifier $n_i$, we have the following notation:
\begin{itemize}
  \item $\SetOfEngines_i$ is the set of engines on the node, which are precisely
  the values of the mapping $\enginemap_i$, and
  \item $\alpha_i$ is a mapping from engine instances into
  their identifiers.
\end{itemize}

\begin{remark}[Modification of nodes and engines]\label{rem:node-operations}
%By design, nodes cannot create, destroy, or modify engines directly.
The creation, destruction, and modification of engines will be \todo{describe in
a phrase} covered in \todo{ref}.
% J:to consider, if an engine is destroyed, is its mailbox engine also destroyed?
% also, what does it implies if an engine is modified? What can be modified any
% way?
These
operations require global coordination through system-level mechanisms rather
than node-local decisions. This architectural constraint exists because nodes
lack awareness of critical system-wide concerns like address registry management,
topology relationships between components, and coordination mechanisms for
distributed state consistency.

%TODO: introduce a forward reference to the related op. rules
\end{remark}

\subsection{System state}\label{sec:state}%
Having defined the concept of nodes, we now define the overall system. A system
state consists of a collection of nodes and a set of in-transit messages that
represent ongoing communication between engines. The system is a variation of a
Kripke structure. We therefore define the possible system states, \emph{the
possible worlds}, and then specify the transition rules that govern its evolution,
one labelled transition at a time.
%TODO: Review this one more time
% When referecing a "system" we are really
% referencing a "system state". It's confusing.

\begin{definition}[State]\label{def:state} A \textbf{state} of the system
$\SystemState$ is a tuple
$\equal{%
    \SystemState^i
}{%
\langle
    \SetOfNodes^i
  , \SetOfMessages^i
  , \SetOfPlugins^i
\rangle
}$ where,
\begin{itemize}
\item $\SetOfNodes^i$ is the set of nodes,
\item $\SetOfMessages^i$ is the set of in-transit messages representing ongoing communications, and
\item $\SetOfPlugins^i$ is the set of system-wide services and functions.
\end{itemize}
\end{definition}

% TODO: what are the implications that  \mu^i is a set?
% for the append rule, we require to (i) access a message in the set, and (ii)
% remove it from the set. However, it's not clear if we should preserve the order
% of the messages in the set. If needed, a timestamp could be added to the messages
% to preserve the order.

\begin{remark}\label{rem:system-time-notation}%
The superscript $i$ in $\SystemState^i$ denotes an instantiation of the system.
This can be viewed as a snapshot of the system at a point in time denoted by
index $i$, with the ability to evolve to a successor state $\SystemState^{i+1}$.
\end{remark}

The \emph{initial state} of the system is denoted by $\SystemState^0$. In principle,
we could define the initial state as %
$\defequal{\SystemState^0}{\systemval{\emptyset}{\emptyset}{\emptyset}}$, %
no nodes, no messages, and no services. Via primitives that we will define
later, we can step from the initial state to a successor state with nodes,
messages, and capabilities.

\begin{remark}\label{rem:system-model}%
The Engine Model \cite{art-engines}, which
inspires the system described here, is a communication-based system model
incorporating temporal aspects and dynamic process creation. Here we retain only
the dynamic process creation capability, leaving temporal considerations for
future work. Like \textsc{DETCs}, we use labelled transition system semantics
where transitions occur when processes react to messages by updating their
state, generating new messages and engines. The associated system in
\textsc{DETC} is defined in terms of active processes and a set of messages,
assuming all messages eventually get delivered. A similar approach is adopted
here for the state in \Cref{def:state}, and we adopt as principle that all
messages are eventually delivered.%, see \Cref{lem:message-delivery}.
\end{remark}

The system must provide primitives to aid in the creation of different
entities such as nodes and engines. The system provides the following functions:

\begin{itemize}
\item \freshidname{} is the function that
generates a fresh identifier for node creation.
% \item $\mathsf{freshName}$ is the function that
% generates a fresh name for an engine.
\item $\typing{\mailboxOfname}{\Address \to \Address}$ is the function that
outputs the address of the dedicated mailbox engine for a given processing
engine.
% \begin{equation*}
% \begin{aligned}
%   &\mailboxOfname : \Address \to \Address\\
%   &\mailboxOfname(\langle n_i, \mathsf{id}_j \rangle) \equiv \langle n_i, \mathsf{id}_j \rangle
% \end{aligned}
% \end{equation*}
% TODO: check if this is used somewhere else
\item $\execEffect$ is the function that executes effects as in \Cref{def:effects}
and outputs the new system state after the effect has been executed.
\end{itemize}

We could also extend the node-specific function $\enginemap_i$ to construct a
global version $\enginemap$ that operates over all engines instances
$\SetOfEngines$ in $\SystemState^i$. Recall $\SetOfEngines_i$ is the collection
of engine instances $\SetOfEngines_i$ for $i$ ranging over the node set. This
allows us to look up addresses for any engine in the system regardless of which
node it runs.

Not described here are the functions for removing, and modifying nodes.
Also, one may want to move engines between nodes. We leave these considerations
for future work.

\section{Operational Semantics}\label{sec:semantics}

The transitions of the associated state machine of the system define both how the
system evolves and how the message-passing mechanism functions.

We first recall a few notation conventions. $\SetOfNodes$ denotes the set of
nodes in the system, $\enginemap$ denotes a mapping from engine addresses to
their corresponding instances. $\SetOfEngines$ is the set of values of
$\enginemap$, that is the set of engine instances in the system. The function
$\alpha$ maps engine instances back to their addresses, while $\SetOfMessages$
represents the set of all in-transit messages.

The operational semantics for the system are defined here
via the reduction relation denoted by $\stepsymbol$, where %
$\sysstep{\SystemState^i}{\SystemState^{i+1}}$ signifies that the system state
$\SystemState^i$ evolves to $\SystemState^{i+1}$. The notation $\SystemState^i$
is mentioned in \Cref{rem:system-time-notation}.

\subsection{System-level initialisation}\label{sec:creation-rules}

%TODO: add a rule for genesis the system state

In the following rules, we assume that $\SystemState$ is of the form %
$(\SetOfNodes, \SetOfMessages, \SetOfPlugins)$, unless otherwise specified.

\begin{definition}[\srule{Node}]\label{rule:s-Node}
\sloppy~\hfill\\[-\baselineskip]%\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
      \defequal{\iota}{\freshid}
\quad %
      \equal{%
        n
      }{%
        \nodeval%
          {\iota}
          {\emptyset}
          {\plugins_n}
      }
$}
\UnaryInfC{$%
\opstep%
{\SystemState}%
{node}%
{%
  ( \SetOfNodes \cup \{n\}
  , \SetOfMessages
  , \SetOfPlugins
  )
}%
$}
\end{prooftree}
\end{definition}

% Rules to introduce engine types and related types.

\subsection{Engine creation}\label{sec:engine-creation-rules}

This section outlines the rules for creating and initializing engines within the
system. The process involves setting up both the processing engine and its
associated mailbox engine, ensuring that each engine is properly configured and
ready to operate.

\begin{definition}[\srule{EngineSpawn}]\label{rule:s-EngineSpawn}
\hspace{-1.5em}
\begin{prooftree}
% Level 1: Node and configuration setup
\AxiomC{$%
\equal{%
  \node{i}
}{%
  \nodeval%
    {\mathsf{id}_i}
    {\enginemap_i}
    {\plugins_i}
}
\in \SetOfNodes
$}
\noLine
\UnaryInfC{$%
\quad \typing{\pi}{\mathsf{Option}(\Address)} %
\quad %
      \equal{%
        \config{\new}%
      }{%
        \configval{\pi}{\process}{\ignore}
      }%
$}
\noLine
\UnaryInfC{$%
\quad \typing{S}{\U} %
\quad \typing{\environment{\new}}{\Env(S)}
$}
\noLine
\UnaryInfC{$%
\quad \typing{\texttt{prot}}{\Msg} %
$}
\noLine
\UnaryInfC{$%
\quad %
      \equal{%
        \engine{\new}
      }{
        \engineval%
          {\texttt{prot}} %
          {b} %
          {\ready{\none}} %
          {\config{\new}} %
          {\environment{\new}}
      }
$}
\noLine
\UnaryInfC{$%
      \defequal{\iota}{\freshid}%
\quad \defequal{\eta}{\freshid} %
\quad %
      \equal{%
        \config{\mailboxOfname}
      }{%
        \configval%
          {\iota} %
          {\mail}
          {\ignore}
      }%
$}
\noLine
\UnaryInfC{$%
\equal{%
  \mailbox{\mathsf{new}}
}{%
  \engineval%
    {\interface{\mailboxOfname}} %
    {\behaviour{\mailboxOfname}} %
    {\status{\mathsf{ready}(\none)}} %
    {\config{\mailboxOfname}} %
    {\environment{\mailboxOfname}}%
  }%
$}
\noLine
\UnaryInfC{$%
\equal{%
  \enginemap'
}{
  \enginemap_i%
    [\iota \mapsto \engine{\mathsf{new}} ]%
    [\eta  \mapsto \mailbox{\mathsf{new}}]
}%
$}
\UnaryInfC{$%
\opstep%
  { \SystemState
  , \node{i}
  , c % the configuration of the engine
  , s % the local environment of the engine
  }
  {spawn}%
  {%
    ( \SetOfNodes[\enginemap_i \mapsto \enginemap']%
    , \SetOfMessages%
    , \SetOfPlugins%
    )
  }%
$}
\end{prooftree}

The terms $\interface{\mailboxOfname}$, $\behaviour{\mailboxOfname}$, and
$\environment{\mailboxOfname}$ refer to the message interface, behaviour, and environment of the mailbox engine, respectively. These follow the specifications outlined in \Cref{def:mailbox-engine}, and default values are used when specific values are not provided.
% TODO: it may seem the function to get mailbox address needs to be defined
% in the node-specific operations.
\end{definition}

\begin{remark}
The rule \sruleref{EngineSpawn} takes place in the context of a node $\node{i}$.
When defining a processing engine $\engine{\mathsf{new}}$, the system also
creates a mailbox engine on the same node, $\mailbox{\mathsf{new}}$. While the
only parameters required by this rule are the processing engine parameters
(configuration and environment) for $\engine{\mathsf{new}}$, one could customize
the mailbox parameters, such as its behaviour $\behaviourtype{\mathsf{new}}$, to
enable specialised message routing or priority-based message organisation.
However, for simplicity, we use a uniform mailbox behaviour across all mailbox
engines.
\end{remark}

\begin{comment}
We could think of adding rules for removing nodes and engines. However,
one question just arises to me: how to handle broken references to those
engines that are removed?
\end{comment}

\subsection{Engine garbage collector}\label{sec:engine-termination}

To prevent \emph{zombie} engine instances---engines in the $\terminated$ state
that could violate liveness and safety guarantees---we need to remove them from
the system. This is accomplished using the \sruleref{Clean} rule.

\begin{definition}[\srule{Clean}]\label{rule:s-Clean}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
  p \in \SetOfEngines%
\quad %
  \propequal{%
    \terminated
  }{%
    \status{p}
  }
$}
\UnaryInfC{$%
\opstep%
  {\SystemState}%
  {clean}%
  {%
    ( \SetOfNodes \setminus \{p\}%
    , \SetOfMessages% TODO: Do we NEED to remove references to e???
    , \SetOfPlugins%
    )
  }%
$}
\end{prooftree}
\end{definition}

\begin{remark}
Notice how the \sruleref{Clean} rule does not remove the engine's mailbox.
Even when removing the processing engine, the mailbox is kept in the system
receiving messages, but not being processed. This ensures that messages are not
lost.
%Also, the mailbox can be reused for future engine instances? as it is a matter
%of changing the engine's configuration.
\end{remark}

\subsection{Message-passing mechanism}\label{sec:message-passing-mechanism}

Our message-passing mechanism aims to support (i) asynchronous communication
with non-blocking sends, (ii) mailbox isolation with guaranteed eventual
delivery, and (iii) explicit lifecycle management for engines.

\begin{definition}[\mrule{Send}]\label{rule:m-Send}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
\engine{s}, \engine{t} \in \SetOfEngines%
\quad %
  \typing{v}{\msgtype{t}}%
\quad %
  \equal{%
    \SystemState
  }{%
    (\SetOfNodes, \SetOfMessages, \SetOfPlugins)
  }%
$}
\UnaryInfC{$%
\opstep%
  {%
    \SystemState, \alpha(\engine{t}), v%
  }
  {send}%
  {%
    ( \SetOfNodes
    , \SetOfMessages \cup%
      \{
        \msgpacket%
          {\alpha(\engine{s})}
          {\mailboxOf{\enginemap(\engine{t})}}
          {v}
      \}%
    , \SetOfPlugins%
    )
  }%
$}
\end{prooftree}
\end{definition}

The \mruleref{Send} rule above supports the sending of messages between engines in
the context of a system state $\SystemState$. When engine $\engine{s}$ sends a message to
engine $\engine{t}$, the rule handles the registration of the message in $\engine{t}$'s
dedicated mailbox. This message must be well-typed according to $\engine{t}$'s message
interface specification, and it is the only way to add a message to $\SetOfMessages$.

%TODO: add a lemma stating that the operational mode
% of an engine found in the destination of a message is
% always an mailbox.

Now, with a message in-transit, the system needs to deliver it to the target engine's
mailbox. This is done via the \mruleref{Enqueue} rule.

\begin{definition}[\mrule{Enqueue}]\label{rule:m-Enqueue}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
\equal{%
  m
}{%
  \msgpacket%
    {\ignore}
    {\mailbox{t}}
    {\append(v)}
} \in \SetOfMessages
\quad %
  \equal{%
    q
  }{%
    \mathsf{env}(\enginemap(\mailbox{t}))
  }%
$}
%
\noLine
\UnaryInfC{$
\equal{%
  \eta
}{%
  \mathsf{fst}(\mailbox{t})
}
\quad %
  \equal{%
    \enginemap'
  }{%
    \enginemap[%
      \mailbox{t}
        \mapsto%
      \engineval%
        {\ignore}
        {\ignore}
        {\ignore}
        {\ignore}
        {q \cup \{v\}}
    ]%
  }
$}
\UnaryInfC{$%
\opstep%
  {\SystemState}
  {enqueue}
  {%
    ( \SetOfNodes[ \enginemap_\eta \mapsto \enginemap' ]
    , \SetOfMessages \setminus \{m\}
    , \SetOfPlugins%
    )
  }
$}
\end{prooftree}
\end{definition}

\begin{definition}[\mrule{Dequeue}]\label{rule:m-Dequeue}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$
  p \in \SetOfMessages%
\quad %
  \equal{
    p
  }{%
    \engineval%
      {\ignore}
      {\ignore}
      {\ready{\ignore}}
      {\ignore}
      {\ignore}
  }
$}
\noLine
\UnaryInfC{$
  \mailbox{p} \in \SetOfEngines%
\quad %
  \equal{
    \mailbox{p}
  }{%
    \mailboxOf{p}
  }
\quad %
  \propequal{%
    \mathsf{env}(\enginemap(\mailbox{p}))
  }{%
   q \cup \{v\}
  }
$}
\noLine
\UnaryInfC{$%
\equal{%
  \SetOfNodes'
}{%
  \SetOfNodes[%
    \enginemap({p})%
      \mapsto%
        \engineval%
          {\ignore}
          {\ignore}
          {\busy{v}}
          {\ignore}
          {\ignore}
    ]%
    [\mailbox{p}
        \mapsto
          \engineval%
          {\ignore}
          {\ignore}
          {\ignore}
          {\ignore}
          {q}
    ]%
}
$}
\UnaryInfC{$%
\opstep%
  {\SystemState}
  {dequeue}
  {
    ( \SetOfNodes'
    , \SetOfMessages
    , \SetOfPlugins
    )
  }
$}
\end{prooftree}
\end{definition}

After a processing engine retrieves a message via the \mruleref{Dequeue}
rule, it enters the \texttt{busy} state to handle the incoming message. This
message processing involves executing the engine's predefined behaviour through
the \sruleref{Process} rule, as defined in~\Cref{sec:engine-processing}.
However, we need first to establish how an engine's behaviour evaluates messages
through its guarded actions, and how the evaluation of those effects changes the
system state.

\subsection{Behaviour evaluation}\label{sec:behavior-eval-rules}

\begin{comment}
restrict attention to machines
whose behaviour is guaranteed to terminate,
\end{comment}

While the system's operational semantics primarily govern inter-engine
communication, individual engines retain full autonomy over the computations
that they perform locally to process messages.
This subsection formalises the evaluation mechanism for
guarded actions that govern engine responses to messages. Specifically, the
following rules define the process by which an engine $p$ consumes a message $v$
from its mailbox, formalised via, first the judgment

\begin{equation}\label{eq:judgement}
\evalstep%
  {p}%
  {v}%
  {E}%
,%
\end{equation}
that is, processing the message $v$ within the context of a busy engine $p$,
results in an effect $E$, and second, a more fine-grained judgment
of the former to introduce the guarded action used to produce the effect.

\begin{equation}\label{eq:guarded-action-eval}
  \guardevalmsgstep%
    {p}
    {g}
    {a}
    {v}
    {E}
  ,
\end{equation}
%
states that in the context of engine $p$, the guarded action%
$\langle g, a \rangle$ produces the effect $E$ when processing the message $v$.

\begin{definition}[\brule{GuardedActionEval}]\label{rule:b-GuardedActionEval}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
  p \in \SetOfEngines%
\quad %
  \typing{v}{\msgtype{p}}%
\quad %
  \equal{%
    p
  }{%
    \engineval%
      {\ignore}
      {\behaviour{p}}
      {\busy{v}}
      {\config{p}}
      {\localenv{p}}
  }
$}
\noLine
\UnaryInfC{$%
  \guardedactionval%
    {g}
    {a}
  \in \behaviour{p}
\quad %
  \propequal{%
    g(v, \config{p}, \localenv{p})
  }{%
    \some{r}
  }
$}
\UnaryInfC{$
\guardevalmsgstep%
  {p}
  {g}
  {a}
  {v}
  {a(r, v, \config{p}, \localenv{p})}
$}
\end{prooftree}
\end{definition}

\begin{definition}[\brule{NoMatchingGuard}]\label{rule:b-NoMatchingGuard}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
  p \in \SetOfEngines%
\quad %
  \equal{%
    p
  }{
    \engineval%
      {\ignore}
      {\behaviour{p}}
      {\busy{\ignore}}
      {\config{p}}
      {\localenv{p}}
  }
$}
\noLine
\UnaryInfC{$%
\guardedactionval{g}{a} \in \behaviour{p}%
\quad %
  \propequal{%
    g(v, \config{p}, \localenv{p})
   }{%
    \none%
   }
$}
\UnaryInfC{$%
\guardevalmsgstep%
  {p}
  {g}
  {a}
  {v}
  {\noop}
$}
\end{prooftree}
\end{definition}

\paragraph{Guard selection strategy}
Behaviours consist of a set of guarded actions. Execution of this behaviour
requires a strategy to select which guarded action to execute. For the sake of
simplicity, we assume that the strategy is deterministic, and that the engine
runs an action only if it is the only one that matches the message. Again, other
guard selection strategies may cause unintended behaviour, and the study of which
strategies are valid, and their semantics is out of scope of this paper.

\begin{definition}[\brule{GuardStrategy}]\label{rule:b-GuardStrategy}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
p \in \SetOfEngines%
\quad %
  \equal{%
    p
  }{%
    \engineval%
      {\ignore}
      {\behaviour{p}}
      {\busy{v}}
      {\ignore}
      {\ignore}
  }
$}
\noLine
\UnaryInfC{$%
  \guardedaction{i} \in \behaviour{p}%
% \quad p , \langle g_i, a_i \rangle \vdash v \downarrow E
\quad %
  \guardevalmsgstep%
    {p}
    {\guard{i}}
    {\action{i}}
    {v}
    {E}
$}
\noLine
\UnaryInfC{$%
\forall_{j\neq i}
  \guardedaction{j} \in \behaviour{p} %
  \,.\
  \guardevalmsgstep%
  {p}
  {\guard{j}}
  {\action{j}}
  {v}
  {\noop}
$}
\UnaryInfC{$%
\evalstep%
  {p}
  {v}
  {E}
$}
\end{prooftree}
\end{definition}

\begin{remark}
Notice that there is no rule here that allows two guards to match the same
message, that is, producing different non-noop effects. This is an example of
unexpected behaviour that a different guard strategy could refine.
\end{remark}

\subsection{Effect execution}\label{sec:effect-execution}

With rules how to get the effect resulting from processing a message, we now
need to state how the system evolves when an effect is executed.

Given a system state $\SystemState$, an effect $E$ and a new system state
$\SystemState'$, we write

\begin{equation}\label{eq:effect-exec}
\effectevalstep%
  {\SystemState}
  {E}%
  {\SystemState'}
\end{equation}

to state that when the effect $E$ is executed in the system state
$\SystemState$, it produces the new system state $\SystemState'$. As the effect
$E$ takes many forms, this judgment is defined by case analysis on the effect
$E$, as shown in the following definitions.


\begin{definition}[\erule{Send}]\label{rule:e-Send}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
  \engine{s}, \engine{t} \in \SetOfEngines%
\quad %
  \equal%
    {
      l_{t} % address
    }{%
      \alpha(\engine{t})
    }
$}
\noLine
\UnaryInfC{$%
      \typing{v}{\msgtype{s}}
\quad \typing{m}{\msgtype{t}}
\quad %
     \evalstep%
        {\engine{s}}
        {v}
        {\send{l_{t}}{m}}
\quad %
      \sendstep%
        {\SystemState}
        {l_{t}}
        {m}
        {\SystemState'}
$}
\UnaryInfC{$
\effectevalstep%
  {\SystemState}
  {\send{l_{t}}{m}}
  {\SystemState'}
$}
\end{prooftree}
\end{definition}

\begin{definition}[\erule{Update}]\label{rule:e-Update}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
      p \in \SetOfEngines%
\quad \typing{S}{\U}%
\quad \typing%
        {
          \environment{p} , s
        }{
          \envtype{p}{S}
        }%
$}
\UnaryInfC{$%
\effectevalstep%
  {(\SetOfNodes, \SetOfMessages, \SetOfPlugins)}
  {\update{s}}
  {(\SetOfNodes[
    \enginemap(\engine{p}) \mapsto %
      \engineval%
        {\ignore}
        {\ignore}
        {\ignore}
        {\ignore}
        {s}
    ]%
   , \SetOfMessages
   , \SetOfPlugins
   )}
$}
\end{prooftree}
\end{definition}


\begin{definition}[\erule{MFilter}]\label{rule:e-MFilter}
  \hspace{-1.5em}
  \begin{prooftree}
  \AxiomC{$%
  \quad p \in \SetOfEngines%
  \quad \typing{f}{\Option{\msgtype{p} \to \Bool}}%
  \quad \evalstep%
            {p}
            {v}
            {\mfilter{f}}%
  $}
  \UnaryInfC{$%
  \effectevalstep%
  {(\SetOfNodes, \SetOfMessages, \SetOfPlugins)}
  {\mfilter{f}}
  {(\SetOfNodes[
    \enginemap(p) 
      \mapsto %
      \engineval%
        {\ignore}
        {\ignore}
        {\ready{f}}
        {\ignore}
        {\ignore}
    ]
    , \SetOfMessages
    , \SetOfPlugins
  )}
$}
  \end{prooftree}
  \end{definition}

\begin{definition}[\erule{Spawn}]\label{rule:e-Spawn}
\hspace{-1.5em}
\begin{prooftree}
% parent engine
\AxiomC{$%
\node{i} \in \SetOfNodes%
\quad p \in \SetOfEngines_i%
\quad %
  \equal%
    {
      c
    }{
      \configval%
        {\alpha(p)}
        {\process}
        {\ignore}
    }
$}
\noLine
\UnaryInfC{$%
      \typing{S}{\U}%
\quad \typing{s}{\envtype{j}{S}}% %FIXME: ENV? which
\quad %
      \spawnstep%
        {\SystemState}
        {\node{i}}
        {c}
        {s}
        {\SystemState'}%
$}
\UnaryInfC{$%
  \effectevalstep%
    {\SystemState}
    {\spawn{c}{s}}
    {\SystemState'}
$}
\end{prooftree}
\end{definition}

\begin{definition}[\erule{Chain}]\label{rule:e-Chain}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
      p \in \SetOfEngines%
\quad v \in \SetOfMessages%
\quad \typing{v}{\msgtype{p}}%
$}
\noLine
\UnaryInfC{$%
     \evalstep%
        {p}
        {v}
        {\chain{\mathit{eff}_1}{\mathit{eff}_2}}%
\quad \effectevalstep%
        {\SystemState}
        {\mathit{eff}_1}
        {\SystemState'}
\quad \effectevalstep%
        {\SystemState'}
        {\mathit{eff}_2}
        {\SystemState''}
$}
\UnaryInfC{$%
\effectevalstep%
  {\SystemState }
  {\chain{\mathit{eff}_1}{\mathit{eff}_2}}
  {\SystemState''}
$}
\end{prooftree}
\end{definition}

\begin{definition}[\erule{Terminate}]\label{rule:e-Terminate}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
      p \in \SetOfEngines%
\quad v \in \SetOfMessages%
\quad \typing{v}{\msgtype{p}}%
\quad \evalstep%
          {p}
          {v}
          {\terminate}%
\quad \opstep%
          {\SystemState}
          {\mathsf{clean}}
          {\SystemState'}
$}
\UnaryInfC{$%
\effectevalstep
  {\SystemState}
  {\terminate}
  {\SystemState'}
$}
\end{prooftree}
\end{definition}

\begin{definition}[\erule{Noop}]\label{rule:e-Noop}
  \hspace{-1.5em}
  \begin{prooftree}
  \AxiomC{$%
        p \in \SetOfEngines%
  \quad v \in \SetOfMessages%
  \quad \typing{v}{\msgtype{p}}%
  \quad \evalstep%
          {p}
          {v}
          {\noop}
  $}
  \UnaryInfC{$%
  \effectevalstep%
    {\SystemState}
    {\noop}
    {\SystemState}
  $}
  \end{prooftree}
\end{definition}

\subsection{Engine processing}\label{sec:engine-processing}

This section introduces the core system-level operational rule
\sruleref{Process} that governs how engines process messages.
Given a system state $\SystemState$, an engine $p$, and a message
$m \in \SetOfMessages$,
the judgment
\begin{equation}\label{eq:process-judgment}
\processstep%
  {\SystemState}
  {p}
  {m}
  {\SystemState'}
\end{equation}
reads as with system state $\SystemState$, when engine $p$ processes the message $m$,
it produces a new system state $\SystemState'$. This rule aims to coordinate three
critical aspects:
\begin{enumerate*}[label=(\roman*)]
  \item effect generation from message processing, and
  \item engine state transitions from \textsf{busy} to \textsf{ready} after successful execution.
\end{enumerate*}

\begin{definition}[\srule{Process}]\label{rule:s-Process}
\hspace{-1.5em}
\begin{prooftree}
\AxiomC{$%
      \node{i} \in \SetOfNodes%
\quad p \in \SetOfEngines_i%
\quad %
      \equal{
        p
      }{
        \engineval%
          {\interface{p}}
          {\behaviour{p}}
          {\busy{m}}
          {\config{p}}
          {\localenv{p}}
      }
$}
\noLine
\UnaryInfC{$%
      \evalstep%
        {p}
        {m}
        {E}
\qquad \effectevalstep%
          {\SystemState}
          {E}
          {(\SetOfNodes', \SetOfMessages', \SetOfPlugins')}
$}
\noLine
\UnaryInfC{$%
\equal{
  \enginemap'
}{
  \enginemap[ %
    \alpha(p) %
      \mapsto %
        \engineval%
          {\interface{p}}
          {\behaviour{p}}
          {\ready{m}}
          {\config{p}}
          {\localenv{p}}
    ]
}
$}
\UnaryInfC{$%
\processstep%
  {\SystemState}
  {p}
  {m}
  {( \SetOfNodes'[%
        \enginemap \mapsto \enginemap'
        ]%
    , \SetOfMessages'
    , \SetOfPlugins'
  )}
$}
\end{prooftree}
\end{definition}

\subsection{System state transition rules}

The inference rules from previous sections collectively govern system state
transitions. We establish the state transition type by showing that system-level
judgments in \Cref{fig:state-transition-rules} induce state transitions:

\begin{equation}
\opstep%
  {\SystemState, \cdots}
  {op}
  {\SystemState'}%
\quad \text{implies}%
\quad \sysstep%
        {\SystemState}
        {\SystemState'}
\end{equation}

\noindent for any system-level atomic operation $\mathsf{op}$.
System-level operations are labelled with identifiers like \srule{Name} in \Cref{fig:state-transition-rules}.

\begin{figure}[h]
  \centering
  \begin{tabular}{ll}
    \textbf{Category} & \textbf{Rules} \\ \hline
    System operations &
        \sruleref{Node},
        \sruleref{EngineSpawn},
        \sruleref{Clean},
        \sruleref{Process} \\
    Message passing &
        \mruleref{Send},
        \mruleref{Enqueue},
        \mruleref{Dequeue} \\
    Behaviour evaluation &
        \bruleref{GuardedActionEval},
        \bruleref{NoMatchingGuard},
        \bruleref{GuardStrategy} \\
    Effect execution &
        \eruleref{Send},
        \eruleref{Update},
        \eruleref{MFilter},
        \eruleref{Spawn},
        \eruleref{Chain},
        \eruleref{Terminate},
        \eruleref{Noop}
  \end{tabular}
  \caption{Rules for the operational semantics categorised by domain.}
  \label{fig:state-transition-rules}
\end{figure}

\begin{remark}
Critical dependencies exist between these rules: \mruleref{Send} enables
inter-engine communication but requires \mruleref{Enqueue} for message
delivery. The \sruleref{Process} rule depends on successful message
consumption via \mruleref{Dequeue} and the related judgment for behaviour
evaluation, as defined in \Cref{sec:behavior-eval-rules}. On the other hand, the
\sruleref{Clean} rule prevents resource leakage by removing terminated
engines. Omission of these components, that is, system-level or message-passing,
would raise undefined behaviour, for example, excluding \mruleref{Enqueue}
would leave messages undelivered, while removing \sruleref{Clean} would
allow infinite accumulation of dormant engines.
\end{remark}

% This should be provable.
% \begin{lemma}[Eventual reliable message delivery]\label{ax:message-delivery}%
%   For any message in transit within
%   system state $\SystemState$, there exists a subsequent state $\SystemState'$ where $m$ is
%   eventually delivered to its intended recipient.
% \end{lemma}

\section{Concluding remarks}\label{sec:related}

The Engine Model~\cite{art-engines}, which serves as our focus within the Actor
model paradigm, shares conceptual ground with Erlang's process-oriented
architecture and similar Actor model implementations that treat actors as
processes ---entities we refer to as
\emph{engines} in our framework.

Fundamentally, engines represent processes augmented with additional structural
components but not particularly crucial to the main design proposed here. In this
paper, we proposed a mailboxes-as-actor approach that naturally extends to any
Actor-like model that shares a process-oriented
paradigm~\cite{taxonomy-of-actor-models}.

\paragraph{Related work}
The Kilim framework~\cite{kilim-framework} aligns with the mailbox decoupling goal
of this paper. It achieves this separation through specialised queues.
However, this implementation treats mailboxes as passive data structures
rather than first-class actors. This limitation persists across most Actor model
implementations like Agha's Classic Actor Model~\cite{foundation97agha}, and
SALSA~\cite{taxonomy-of-actor-models}, where mailboxes typically function as
internal components with limited exposed functionality beyond basic queue
operations, or like Erlang, where mailbox interaction remains a low-level system
primitive with opaque handling~\cite{erlang-2010}.

In short, by promoting mailboxes to opaque first-class actors, we combine
Kilim's structural separation with Erlang's process-oriented operational
transparency, plus the ability to implement custom mailbox behaviours.
These mailbox engines act as \emph{secretaries} ---intelligent brokers that can
filter content, prioritize messages, delay delivery, or debug issues through
configurable rules.

\paragraph{Multiple mailboxes per processing actor}
While our architecture could naturally accommodate multiple mailbox actors per
processing actor, a pattern that might enable specialised message routing or
priority queues, we consciously omit this extension from our core design. This
restriction serves to preserve the system's verifiable properties. Concurrent
pulling from multiple mailboxes could introduce race conditions that violate our
exactly-once delivery guarantee. The coordination overhead required for atomic
operations across distributed mailboxes would also compromise the simplicity
benefits of our decoupled architecture. Therefore, we intentionally maintain a
one-to-one relationship between processing actors and their mailboxes, as this
constraint ensures reliable message ordering while preserving the opacity and
encapsulation benefits central to our design.

\paragraph{System properties}
There is nothing that prevents us from proving system properties for this
extended model. A possible starting point would be to add a timeout primitive to
the engine model and demonstrate that the system remains live and makes progress
even in the presence of timeouts for practical implementations.

\paragraph{Practical implementations}
While several systems implement related concepts, no existing implementations
fully adopt the mailbox-as-actor design. Actor frameworks like Kilim for
Java~\cite{kilim-framework} demonstrate partial alignment by decoupling
mailboxes from actors ---treating them as object/work-queues where actors pull
messages when ready to process them. We identify analogous approaches in the
JavaScript ecosystem, such as the WeChaty library employed in the industrial chatbot
applications using actor-like architectures~\cite{wechaty}.
Our work's distinct contribution
lies in formally promoting mailboxes to first-class citizens, actors, within the
model itself, rather than treating them as implementation details or external
components.

\paragraph{Rules as way to guide implementation}
The operational semantics given in \Cref{sec:semantics} can be seen as a way to
guide the implementation of the \say{mailbox-as-an-actor} approach, described in this
paper. We presented the rules in such a way that they should be easy to reason about and
implement.

The system-level rules are given in \Cref{sec:creation-rules}. In this section,
we described how the system grows from the initial empty state, with the addition
of nodes, engines, and later on messages in \Cref{sec:message-passing-mechanism}. We
cover other rules such as \sruleref{Clean} as a garbage-collector for engines,
however, we did not cover rules for on-demand engine or node
deletion/disconnection. This can be seen as an extension of the model.

Our message-passing mechanism implements the rules given in
\Cref{sec:message-passing-mechanism}. The \mruleref{Send} rule enables
asynchronous message realisation in the system, allowing concurrent message
generation without sender blocking. Messages are reliably delivered to
mailbox engines through the \mruleref{Enqueue} rule, which implements atomic
\say{enqueuing} operations. The \mruleref{Dequeue} rule manages message
consumption from mailboxes.

Finally, and probably one of the most interesting rules, the \sruleref{Process}
rule enables atomic state transitions by combining behaviour evaluation, rules
in \Cref{sec:behavior-eval-rules} with effect handling, and rules in
\Cref{sec:effect-execution}. Together, the rules in \Cref{sec:semantics} form a
complete system for reliable asynchronous communication with proper resource
management.

\paragraph{Future work}
As mentioned in \Cref{sec:related}, we plan to implement a prototype in Elixir
leveraging its Erlang-compatible actor model support on the BEAM virtual
machine.%
% The prototype source code is available in the Anoma Foundation's public
% repository at \url{https://github.com/anoma/engine}.

\section{Acknowledgements}

This work was inspired by discussions  with Jeremy Ornelas and Christophe De
Troyer.

\bibliography{art.bib,ref.bib}
\appendix

% \section{Key-value guarded actions}

% \Cref{fig:keyvalue-engine-get} shows the guarded actions for the key-value store engine.

% The guard $g_{\mathsf{get}}$ checks if the message is a $\msg{Get}$ request and extracts the key.
% The action $a_{\mathsf{get}}$ retrieves the value for the key, increments the access count,
% and sends the result back to the sender.
% \begin{align*}
% &\typing{
%   g_{\mathsf{put}}
% }{
%   \In{\mathsf{kv}} \to \Option{\mathsf{Key} \times \mathsf{Val}}
%   \,.
% } \\
% &\defequal{g_{\mathsf{put}}((m, c, \ignore))}{
%   \mathsf{case}\ m\ \mathsf{of}\ \{\\
%   &\hspace{3mm}  \msg{Put}(k, v) \Rightarrow 
%         \mathsf{if}\ \neg\, \mathsf{readOnly}(c)\ 
%         \mathsf{then}\ \some{\langle k, v \rangle}\ 
%         \mathsf{else}\ \none;\\
%   &\hspace{3mm} \ignore \Rightarrow \none
%   \}
%   \,.
% }\\[2mm]
% &\typing{a_{\mathsf{put}}}{
%   (\mathsf{Key} \times \mathsf{Val}) \times \In{\mathsf{kv}} \to 
%   \effecttype{\mathsf{kv}}{S_{\mathsf{kv}}}
% } \\
% &\defequal{a_{\mathsf{put}}((k, v), (m, c, (\mathsf{st}, \mathsf{cnt})))}{
%   \chain{
%     \update{(
%       \mathsf{st}[k \mapsto v],
%       \mathsf{cnt}
%     )}
%   }{
%     \send{\sender{m}}{\msg{Result}(\some{v})}
%   }
% }
% \end{align*}

% The guard $g_{\mathsf{put}}$ checks if the message is a $\msg{Put}$ request and verifies the store
% is not in read-only mode. The action $a_{\mathsf{put}}$ updates the store with the new key-value pair
% and confirms the operation to the sender.

% \begin{align*}
% &\typing{
%   g_{\mathsf{delete}}
% }{
%   \In{\mathsf{kv}} \to \Option{\mathsf{Key}} \,.
% } \\
% &\defequal{g_{\mathsf{delete}}(m, c, \ignore)}{
%   \mathsf{case}\ m\ \mathsf{of}\ \{\\
%     &\msg{Delete}(k) \Rightarrow\\
%     &\quad\mathsf{if}\ \neg \mathsf{readOnly}(c)\ \mathsf{then}\ \some{k}\ \mathsf{else}\ \none;\\
%     &\ignore \Rightarrow \none\\
%   \}
% }\\[2mm]
% &\typing{a_{\mathsf{delete}}}{
%   \mathsf{Key} \times \In{\mathsf{kv}} \to 
%   \effecttype{\mathsf{kv}}{S_{\mathsf{kv}}}
% } \\
% &\defequal{a_{\mathsf{delete}}(k, m, c, e)}{
%   \begin{array}{l}
%     \mathsf{let}\ v = \mathsf{store}(e).get(k)\ \mathsf{in} \\
%     \mathsf{let}\ newStore = \mathsf{store}(e).remove(k)\ \mathsf{in} \\
%     \chain{
%       \update{e[\mathsf{store} \mapsto newStore]}
%     }{
%       \send{m.\mathsf{sender}}{\msg{Result}(v)}
%     }
%   \end{array}
% }
% \end{align*}

% The guard $g_{\mathsf{delete}}$ checks if the message is a $\msg{Delete}$ request and ensures
% the store is not in read-only mode. The action $a_{\mathsf{delete}}$ retrieves the current value,
% removes the key from the store, and returns the previous value to the sender.

% \begin{align*}
% &\typing{g_{\mathsf{stats}}}{\In{\mathsf{kv}} \to \Option{\mathsf{Key}}} \\
% &\defequal{g_{\mathsf{stats}}(m, \ignore, \ignore)}{
%   \mathsf{case}\ m\ \mathsf{of}\ \{\\
%     &\msg{GetStats}(k) \Rightarrow \some{k};\\
%     &\ignore \Rightarrow \none\\
%   \}
% }\\[2mm]
% &\typing{a_{\mathsf{stats}}}{
%   \mathsf{Key} \times \In{\mathsf{kv}} \to 
%   \effecttype{\mathsf{kv}}{S_{\mathsf{kv}}}
% } \\
% &\defequal{a_{\mathsf{stats}}(k, m, c, e)}{
%   \begin{array}{l}
%     \mathsf{let}\ cnt = \mathsf{accessCount}(e).getOrDefault(k, 0)\ \mathsf{in} \\
%     \mathsf{let}\ exists = \mathsf{store}(e).containsKey(k)\ \mathsf{in} \\
%     \send{m.\mathsf{sender}}{\msg{StatsResult}(k, cnt, exists)}
%   \end{array}
% }
% \end{align*}

% The guard $g_{\mathsf{stats}}$ checks if the message is a $\msg{GetStats}$ request and extracts the key.
% The action $a_{\mathsf{stats}}$ retrieves the access count and checks if the key exists in the store,
% then sends these statistics back to the sender.

% The key-value store's behaviour consists of guarded actions 
% $\langle g_{\mathsf{get}}, a_{\mathsf{get}} \rangle$, 
% $\langle g_{\mathsf{put}}, a_{\mathsf{put}} \rangle$, 
% $\langle g_{\mathsf{delete}}, a_{\mathsf{delete}} \rangle$, and
% $\langle g_{\mathsf{stats}}, a_{\mathsf{stats}} \rangle$.

% Why not just one guarded action? Using multiple guarded actions provides clearer separation of concerns, better maintainability, and more efficient message handling than combining all logic into a single complex guard-action pair.



\end{document}
