
% Title
\newcommand{\pubtitle}{Mailbox Actors}

\newcommand{\pubauthA}{Jonathan Prieto-Cubides}
\newcommand{\pubaffilA}{a}
\newcommand{\orcidA}{0000-0002-8449-3812}
\newcommand{\authemailA}{jonathan@heliax.dev}
% \newcommand{\eqcontribA}{}

\newcommand{\pubauthB}{Anthony Hart}
\newcommand{\pubaffilB}{a}
\newcommand{\orcidB}{0009-0004-4116-5538}
\newcommand{\authemailB}{anthony@heliax.dev}
\newcommand{\eqcontribB}{}

\newcommand{\pubauthC}{Tobias Heindel}
\newcommand{\pubaffilC}{a}
\newcommand{\orcidC}{0000-0003-3371-8564}
\newcommand{\authemailC}{tobias@heliax.com}

% Institutions/Affiliations
\newcommand{\pubaddrA}{Heliax AG}

% Corresponding author mail
\newcommand{\pubemail}{\authemailA}

\newcommand{\pubabstract}{%
%
We present a formal framework for actor systems in which mailboxes are promoted
to independent, first-class actors---native actors with their own
configuration, behaviour, and lifecycle. In traditional actor models, mailboxes
are opaque runtime primitives or passive data structures, limiting formal
reasoning about message delivery policies. Our design decouples message storage
from processing logic by automatically pairing each processing actor with a
dedicated mailbox actor whose guarded-action behaviour governs delivery,
filtering, and scheduling. This separation enables modular, replaceable mailbox
implementations and independent scaling of message ingestion and processing,
while adding no configuration burden for simple engines that need only the
default first-in-first-out policy.

We formalise the system using a predicative Martin-L\"of type theory with
dependent types. All type families are indexed by a finite type of engine type
indices, ensuring type-safe communication by construction. We provide a
complete operational semantics as a labelled transition system, covering engine
creation with automatic mailbox pairing, message passing, behaviour evaluation,
effect execution, and engine lifecycle management. We establish type
preservation, progress, effect determinism under non-overlapping guards, and
mailbox isolation as metatheoretic properties.
}


% Any keywords to be displayed under the abstract
\keywords{
Actor Model \sep
Message-Passing Systems \sep
Mailbox Actors \sep
Operational Semantics \sep
Dependent Types \sep
Formal Methods
}

% Supplementary space between title/abstract and text, if needed
% \newcommand{\pubVadj}{0pt}

% ! DO NOT REMOVE OR MODIFY !
\input{aux-preamble.tex}
% The preprint DOI to be used as a link in the paper
\pubdoi{10.5281/zenodo.14915469}
% \history{(Received Month DD, YYYY; Revised Month DD, YYYY ; Published: August 21, 2023)}
\pubdate{\today}
